Modal Logics in the Coq Proof Assistant

### 2014-12-08 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Add abstract and introduction

\usepackage{url,amsmath,amssymb,fancyvrb,color}

\newcommand{\imp}{\rightarrow}

@@

\title{
God in Coq: \\
Modal Logic in Coq \\ 
and Scott's Version of G\"{o}del's Proof of God's Existence \\
{\small (\emph{Rough Diamond Proof Pearl})}
%   \thanks{
%   This work has been supported by the 
%   German Research Foundation under grant
%   BE2501/9-1.}
}

@@

@@

G\"{o}del's proof of God's existence is challenging to formalize and
verify because it requires an expressive logical language with modal
operators ($\Diamond$ and $\Box$) and higher-order quantifiers.
Specialized modal logic tools lack higher-order capabilities, while
higher-order provers lack built-in support for modalities.  However,
by embedding modal logics into higher-order logic, it is in principle
possible to use standard higher-order automated and interactive
theorem provers for reasoning about and within modal logics
\cite{J23,B9}.  To realize this potential, formalizations
\cite{FormalTheologyRepository} of G\"odel's proof in TPTP THF
\cite{J22}, Coq \cite{Coq} and Isabelle \cite{Isabelle} have been
completed so far, in this order. This paper focuses on the
formalization in Coq. The implementation of the embedding is discussed
in detail in Section \ref{sec:modal} and Scott's version of G\"odel's
proof is shown in Section \ref{sec:proof}.

@@

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}. The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type that depends on possible worlds. Possible worlds are connected by an accessibility relation.

@@

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\Parameter r: i -> i -> \Prop. \com{(* Accessibility relation for worlds *)}
\end{Verbatim}
\end{small}

@@

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem dia_box_to_box: [ mforall p, (dia (box p)) m-> (box p) ].
\Proof. mv. intro p. intro H1. dia_e H1. box_i. box_e H1 H2. exact H2. eapply transitivity.
  apply symmetry. exact R.
  exact R0.
\Qed.
\end{Verbatim}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  
G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). The formalization shown in this section aims at
being as similar as possible to Dana Scott's version of the proof \cite{ScottNotes}. The formulation and numbering of axioms, definitions and theorems is the same as in Scott's notes. Even the Coq proof scripts follow all the steps in Scott's notes. Scott's assertions are emphasized with comments. In contrast to the formalization in Isabelle \cite{Archiv}, where automation via Metis \cite{Hurd03first-orderproof} and Sledgehammer \cite{Sledgehammer} using tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II} and Satallax \cite{Satallax} has been successfully employed, the formalization in Coq used no automation. This was a deliberate choice, mainly because it allowed a qualitative evaluation of the convenience of the possible world semantic embedding approach for \emph{interactive} theorem proving. Moreover, in order to formalize precisely Scott's version and not another automatically found version, automation would have to be heavily limited anyway. Furthermore, the deliberate preference for simple tactics (mostly \emph{intro}, \emph{apply} and the modal tactics described in the previous section) results in proof scripts that could be read as natural deduction proofs. This hopefully makes the formalization more accessible to those who are not experts in Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\com{(* Axiom A1: either a property or its negation is positive, but not both *)}
\Axiom axiom1a : [ mforall p, (Positive (\fun x: u => m~(p x))) m-> (m~ (Positive p)) ].
\Axiom axiom1b : [ mforall p, (m~ (Positive p)) m-> (Positive (\fun x: u => m~ (p x))) ].

\com{(* Axiom A2: a property necessarily implied by a positive property is positive *)}
\Axiom axiom2: [ mforall p, mforall q, 
  Positive p m/\bslash (box (mforall x, (p x) m-> (q x) )) m-> Positive q ].

@@

\com{(* Definition D1: God: a God-like being possesses all positive properties *)}
\Definition G(x: u) := mforall p, (Positive p) m-> (p x).

@@

\com{(* Definition D2: essence: an essence of an individual is a property possessed by it} 
\com{and necessarily implying any of its properties *)}
\Definition Essence(p: u -> o)(x: u) := (p x) m/\bslash mforall q, 
                                        ((q x) m-> box (mforall y, (p y) m-> (q y))).
\Notation "p 'ess' x" := (Essence p x) (at level 69).

\com{(* Theorem T2: being God-like is an essence of any God-like being *)}
\Theorem theorem2: [ mforall x, (G x) m-> (G ess x) ].
\Proof. mv. intro g. intro H1. unfold Essence. split.
  exact H1.
  intro q. intro H2. assert (H3: ((Positive q) w)).
    proof_by_contradiction H4. unfold G in H1. apply axiom1b in H4. apply H1 in H4. 
                                                                    contradiction. 
    cut (box (Positive q) w). \com{(* Scott *)}
      apply K. box_i. intro H5. intro y. intro H6. unfold G in H6. apply (H6 q). exact H5.
      apply axiom4. exact H3.
\Qed.

\com{(* Definition D3: necessary existence: necessary existence of an individual }
\com{is the necessary exemplification of all its essences *)}
\Definition NE(x: u) := mforall p, (p ess x) m-> box (mexists y, (p y)).

@@

The successful formalization of Scott's version of G\"odel's proof of
God's existence indicates that the \emph{embedding} of modal logics
into higher-order logics via the \emph{possible world semantics} is a
viable approach for interactive theorem proving within modal logics.
The minimalistic implementation of the embedding (Section
\ref{sec:modal}) takes special care to hide the underlying possible
world machinery from the user.  An inspection of the proof scripts in
Section \ref{sec:proof} shows that this goal has been achieved. The
user does not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for him.
Moreover, for subgoals that do not involve modalities, the user has all the
usual interactive tactics at his/her disposal.
%The implementation of hints to allow automation tactics to fully cope with the embedding remains for future work. Nevertheless, at this stage a tactic such as \texttt{firstorder} is able to prove all lemmas and theorems in Section \ref{sec:modal} fully automatically.

The theological implications of the verified correctness of this proof
certainly depend on a critical discussion of the underlying concepts,
definitions and axioms, which is beyond the scope of this paper. What
can be said though is that our work presents further evidence that
theism is at least not (obviously) irrational.

Related work includes formalizations of Anselm's argument by Rushby
\cite{Rushby}, and by Oppenheimer and Zalta \cite{Zalta}.  Clearly,
the application of theorem proving technology to the domain of
theoretical philosophy can --- as already pictured by Leibniz ---
be very fruitful for both areas. In fact, new results of
philosophical interest have been obtained by our
work~\cite{arXiv}. Moreover a first, basic infrastructure for
interactive and automated reasoning in higher-order modal logics has
been created and will be further improved in the future.

\paragraph{Acknowledgements:} Thanks to Cedric Auger and Laurent Th\'ery, for their answers to our newbie questions about Ltac in the Coq-Club mailing-list.

@@

\bibitem{arXiv}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock G{\"o}del's God in Isabelle/HOL. 
\newblock {\em Archive of Formal Proofs}, 2013.

>>>>

\usepackage{fancybox} % Must be before "fancyvrb"
\usepackage[utf8]{inputenc}
\usepackage{url,amsmath,amssymb,fancyvrb,color}
\usepackage{proof}
\usepackage{xspace}

\usepackage{calculi}

\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\all}{\forall}
\newcommand{\ex}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly

@@

\newcommand{\Coq}{\texttt{Coq}\xspace}
\newcommand{\Isabelle}{\texttt{Isabelle}\xspace}

\title{
Modal Logics in the Coq Proof Assistant
}

@@

\begin{abstract} 
This paper describes an embedding of higher-order modal
logics in the \Coq proof assistant. {\Coq}'s capabilities are thus extended in a
minimalistic manner, which is nevertheless sufficient for the formalization
of significant modal proofs. The elegance and convenience of this approach, 
from a user perspective, are illustrated here with the successful formalization of Gödel's ontological argument. 
\end{abstract}

@@

Modal logics extend the usual formal logic language with modal operators ($\Box$ and $\Diamond$) and are characterized by the \emph{necessitation rule}, according to which $\Box A$ is a theorem if $A$ is a theorem, although $A \imp \Box A$ is not necessarily a theorem. Various informal concepts, such as \emph{necessity and possibility}, \emph{knowledge and belief}, \emph and \emph{temporal globality and eventuality}, have been formalized with the help of modal operators. Therefore, modal logics \cite{ModalLogic} are well-established for a wide variety of application domains, ranging from philosophy \cite{ToDo} to formal software verification \cite{ToDo:TLAPlus}.

However, general automated reasoning support for modal logics is still not as well-developed as for classical logics. Current deduction tools for modal logics are often limited to propositional, quantifier-free, fragments \cite{ToDo:SeeBjornsProposal} or tailored to particular modal logics and their applications \cite{ToDo:StephanMerz}. Recently, first-order automated deduction techniques based on tableaux, sequent calculi and connection calculi have been generalized to cope with modalities directly and a few new provers have been implemented \cite{JensOtten}.

Another recently explored possibility \cite{J23,B9,J22} is the embedding of first-order and higher-order modal logics into classical higher-order logics, for which existing higher-order automated theorem provers \cite{ToDo:Leo,Satallax} exist.
Embedding is flexible, because various modal logics can be easily supported by adding their characteristic axioms. Furthermore, it is relatively simple to implement, because it does not require any modification in the source code of the higher-order prover. The prover can be used as is, and only the input files provided to the prover must be especially encoded. Furthermore, the efficacy and efficiency of the embedding approach has been confirmed in philosophically interesting benchmarks \cite{FormalTheologyRepository}. These qualities make embedding a convenient approach for \emph{fully automated} reasoning.

However, one may wonder whether the embedding approach is adequate for \emph{interactive} reasoning, when the user proves theorems by interacting with a proof assistant. The main goal of this paper is to answer this question positively. The \Coq proof assistant was chosen because of the authors' greater familiarity with the tactic language of this system.

One major initial concern was whether the embedding could be a disturbance to the user.
Fortunately, by using \Coq's \texttt{Ltac} tactic language, we were able to  define intuitive new tactics that hide the technical details of the embedding from the user. The resulting infra-structure for modal reasoning within \Coq (as described in Section \ref{sec:modal}) provides a user experience where modalities can be handled transparently and straightforwardly. Therefore, a user with basic knowledge of modal logics and \Coq's tactics should be able to use (and extend) our implementation with no excessive overhead.

In order to illustrate the use of the implemented embedding, we show here the formalization of Scott's version \cite{Scott} of Gödel's ontological argument for God's existence (in Section \ref{sec:proof}). This proof was chosen mainly for two reasons. Firstly, it requires not only modal operators, but also higher-order quantification. Therefore, it is beyond reach for specialized propositional and first-order theorem provers. Secondly, this argument addresses an ancient problem in Philosophy and Metaphysics, which has nevertheless received a lot of attention in the last 15 years, because of the discovery of the modal collapse \cite{ToDo}. This proof lies in the center of a vast and largely unexplored application domain for automated and interactive theorem provers.

The simpler ontological argument of Anselm has been automatically verified with \texttt{PVS} by Rushby \cite{Rushby} and with first-order theorem by Oppenheimer and Zalta \cite{Zalta}. Gödel's argument was automatically verified in our previous work on embedding-based fully automated modal theorem proving \cite{ArXiv,AFP}. But this paper presents the first fully interactive and detailed formalization of this proof in a proof assistant.

@@

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}.

The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type \texttt{o} that depends on possible worlds:

@@

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\end{Verbatim}
\end{small}

\noindent
Possible worlds are connected by an accessibility relation, 
which can be represented in \Coq by a parameter \texttt{r}, as follows:

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter r: i -> i -> \Prop. \com{(* Accessibility relation for worlds *)}
\end{Verbatim}
\end{small}

@@

\newcommand{\s}{\qquad}

\begin{calculus}
{propositional rules}
{fig:PropositionalRules}

\vspace{1em}

\s\s
\infer[\bot_E]{A}{ \bot }
\s
\infer[\imp_I]{A \imp B}{ B }
\s
\infer[\imp_I^n]{A \imp B}{ \infer*{B}{\infer[n]{A}{}} }
\s
\infer[\imp_E]{B}{A & A \imp B}

\vspace{2em}

\s\s
\infer[\wedge_I]{A \wedge B}{A & B}
\s\s
\infer[\wedge_{E_1}]{A}{A \wedge B}
\s\s
\infer[\wedge_{E_2}]{B}{A \wedge B}

% \vspace{2em}

% \s\s
% \infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
% \s\s
% \infer[\vee_{I_1}]{A \vee B}{A}
% \s\s
% \infer[\vee_{I_2}]{A \vee B}{B}

\vspace{1em}
\end{calculus}

\begin{calculus}
{double negation elimination}
{fig:Classical}
\infer[\neg\neg_E]{A}{ \neg\neg A }
\end{calculus}

\begin{calculus}
{quantifier rules}
{fig:QuantifierRules}

\vspace{1em}

\s
\infer[\all_I]{\all x_{\tau}. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x_{\tau}. A[x] }
\s\s
\infer[\ex_I]{\ex x_{\tau}. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\beta]}{ \ex x_{\tau}. A[x] }

\vspace{1em}

\begin{center}
\textbf{eigen-variable conditions:} \\
if $\rho$ is a $\all_I$ inference eliminating a variable $\alpha$, then any occurrence of $\alpha$ in the proof should be an ancestor of the occurrence of $\alpha$ eliminated by $\rho$; \\
if $\rho$ is a $\ex_E$ inference introducing a variable $\beta$, then any occurrence of $\beta$ in the proof should be a descendant of the occurrence of $\beta$ introduced by $\rho$.
\end{center}

\vspace{1em}

\end{calculus}

\begin{calculus}
{Rules for Modal Operators}
{fig:NDK}

\vspace{1em}

\s\s\s\s
\infer[\nec_I]{\nec A}{\omega: \fbox{\infer*{A}{}} }
\s\s\s\s
\infer[\nec_E]{w: \fbox{ \infer*{}{A} } }{\nec A}

\vspace{2em}

\s\s\s\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s\s\s\s\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

\vspace{1em}

\begin{center}
\textbf{eigen-box condition:}\\ 
$\omega$ must be a fresh name for a box. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are made.
\end{center}

\vspace{1em}

\end{calculus}

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem dia_box_to_box: [ mforall p, (dia (box p)) m-> (box p) ].
\Proof. mv. 
intros p H1. dia_e H1. box_i. box_e H1 H2. exact H2. eapply transitivity. 
  apply symmetry. exact R. 
  exact R0.
\Qed.
\end{Verbatim}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  
G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). The formalization shown in this section aims at
being as similar as possible to Dana Scott's version of the proof \cite{ScottNotes}. The formulation and numbering of axioms, definitions and theorems is the same as in Scott's notes. Even the Coq proof scripts follow all the steps in Scott's notes. Scott's assertions are emphasized with comments. In contrast to the formalization in Isabelle \cite{AFP}, where automation via Metis \cite{Hurd03first-orderproof} and Sledgehammer \cite{Sledgehammer} using tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II} and Satallax \cite{Satallax} has been successfully employed, the formalization in Coq used no automation. This was a deliberate choice, mainly because it allowed a qualitative evaluation of the convenience of the possible world semantic embedding approach for \emph{interactive} theorem proving. Moreover, in order to formalize precisely Scott's version and not another automatically found version, automation would have to be heavily limited anyway. Furthermore, the deliberate preference for simple tactics (mostly \emph{intro}, \emph{apply} and the modal tactics described in the previous section) results in proof scripts that could be read as natural deduction proofs. This hopefully makes the formalization more accessible to those who are not experts in Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\com{(* Axiom A1:} 
\com{   either a property or its negation is positive, but not both *)}
\Axiom axiom1a : 
  [ mforall p, (Positive (\fun x: u => m~(p x))) m-> (m~ (Positive p)) ].

\Axiom axiom1b : 
  [ mforall p, (m~ (Positive p)) m-> (Positive (\fun x: u => m~ (p x))) ].

\com{(* Axiom A2: }
\com{   a property necessarily implied by a positive property is positive *)}
\Axiom axiom2: [ mforall p, mforall q, 
  Positive p m/\bslash (box (mforall x, (p x) m-> (q x) )) m-> Positive q ].

@@

\com{(* Definition D1: }
\com{   God: a God-like being possesses all positive properties *)}
\Definition G(x: u) := mforall p, (Positive p) m-> (p x).

@@

\com{(* Definition D2: }
\com{   essence: an essence of an individual is a property possessed by it} 
\com{   and necessarily implying any of its properties *)}
\Definition Essence(p: u -> o)(x: u) := 
    (p x) m/\bslash mforall q, ((q x) m-> box (mforall y, (p y) m-> (q y))).
\Notation "p 'ess' x" := (Essence p x) (at level 69).

\com{(* Theorem T2: being God-like is an essence of any God-like being *)}
\Theorem theorem2: [ mforall x, (G x) m-> (G ess x) ].
\Proof. mv. intro g. intro H1. unfold Essence. split.
  exact H1.
  intro q. intro H2. assert (H3: ((Positive q) w)).
    proof_by_contradiction H4. unfold G in H1. apply axiom1b in H4. 
    apply H1 in H4. contradiction.

    cut (box (Positive q) w). \com{(* Scott *)}
      apply K. box_i. intro H5. intro y. intro H6. 
      unfold G in H6. apply (H6 q). exact H5.

      apply axiom4. exact H3.
\Qed.

\com{(* Definition D3: }
\com{   necessary existence: necessary existence of an individual }
\com{   is the necessary exemplification of all its essences *)}
\Definition NE(x: u) := mforall p, (p ess x) m-> box (mexists y, (p y)).

@@

The successful formalization of Scott's version of G\"odel's proof of
God's existence indicates that the \emph{embedding} of modal logics
into higher-order logics via the \emph{possible world semantics} is a
viable approach for interactive theorem proving within modal logics.
The minimalistic implementation of the embedding (Section
\ref{sec:modal}) takes special care to hide the underlying possible
world machinery from the user.  An inspection of the proof scripts in
Section \ref{sec:proof} shows that this goal has been achieved. The
user does not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for him/her.
Moreover, for subgoals that do not involve modalities, the user has all the
usual interactive tactics at his/her disposal.
%The implementation of hints to allow automation tactics to fully cope with the embedding remains for future work. Nevertheless, at this stage a tactic such as \texttt{firstorder} is able to prove all lemmas and theorems in Section \ref{sec:modal} fully automatically.

HOL Light, Isabelle

The theological implications of the verified correctness of this proof
certainly depend on a critical discussion of the underlying concepts,
definitions and axioms, which is beyond the scope of this paper. 
%What
%can be said though is that our work presents further evidence that
%theism is at least not (obviously) irrational. 
Clearly, the application of theorem proving technology to the domain of
theoretical philosophy can --- as already pictured by Leibniz ---
be very fruitful for both areas. In fact, new results of
philosophical interest have been obtained by our
work~\cite{arXiv}. Moreover a first, basic infrastructure for
interactive and automated reasoning in higher-order modal logics has
been created and will be further improved in the future.

\paragraph{Acknowledgements:} we thanks Cedric Auger and Laurent Th\'ery, for their answers to our questions about Ltac in the Coq-Club mailing-list.

@@

\bibitem{AFP}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock G{\"o}del's God in Isabelle/HOL. 
\newblock {\em Archive of Formal Proofs}, 2013.
